Global Deaths Due to Air Pollution

### 2022-05-03 outline.Rmd
changed high and low ordering

Import the deaths-due-to-air-pollution data
```{r}
deaths_df <- data.frame(read.csv("death-rates-from-air-pollution.csv"))
glimpse(deaths_df)
```

@@

```{r, echo=FALSE}
#selecting high-population countries from the world population data frame 
high_pop_countries <- world_pop %>% 
  filter(Year> 1995 & Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
  group_by(Year)

high<- ggplot(high_pop_countries, aes(x=Country.Name, y=Count, color=Country.Name))+
  geom_boxplot()+
  labs(title="Selection of high-population countries", x="Country", y="Population (in millions)")
ggplotly(high)

#selecting low-population countries from the world population data frame 
low_pop_countries <- world_pop %>% 
  filter(Year> 1995 & Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand'))%>% 
  group_by(Year)

low<- ggplot(low_pop_countries, aes(x=Country.Name, y=Count, color=Country.Name))+
  geom_boxplot()+
  labs(title="Selection of low-population countries", x="Country", y="Population (in millions)")
ggplotly(low)

@@

Next, we are going to see the death count for high and low populated countries using the deaths dataframe.
```{r, echo=FALSE}
#selecting high-population deaths from death dataframe 
high_pop_death <- deaths_df %>% 
  filter(year > 1995 & country %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
  group_by(year)

#selecting low-population deaths from death dataframe 
low_pop_death <- deaths_df %>% 
  filter(year > 1995 & country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand')) %>% 
  group_by(year)

@@

# Death Count
> Which country has the highest average death count?

@@

```{r, echo=FALSE, message=FALSE}
hp_countries_population <- world_pop %>% 
  filter(Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia'), Year > 1995) %>% 
  group_by(Country.Name) %>% 
  select(Count) %>% 
  summarize(average_population = mean(Count))

lp_countries_population <- world_pop %>% 
  filter(Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand'), Year > 1995) %>% 
  group_by(Country.Name) %>% 
  select(Count) %>% 
  summarize(average_population = mean(Count))

@@

#Low Population Pollutant Averages 
low_poll <- deaths_df %>% 
  group_by(country) %>% 
  filter(country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand')) %>% 
  select(country, indoor_deaths, outdoor_deaths, ozone_deaths) %>% 
  summarize(avg_indoor = mean(indoor_deaths), avg_outdoor = mean(outdoor_deaths), avg_ozone = mean(ozone_deaths))
kable(low_poll)
```

@@

# Pollution Over Time 
> Let's look at the previous two decades and compare the death count

> has there been a change?

@@

death_06_low<- deaths_df %>% 
  group_by(country) %>% 
  filter(year == 2006, country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'Tonga')) %>%
  select(total_deaths) %>% 
  summarize(Low_Deaths_06 = mean(total_deaths))

@@

The first decade 1996-2006.
```{r, echo=FALSE}
#High Population Deaths 1996-2006 
p_high_first <- all_high_first %>%
  group_by(country) %>% 
  ggplot(aes(x= High_Deaths_96, y= High_Deaths_01, color= High_Deaths_06)) +
  geom_point() +
  facet_wrap(~country)

@@

This shows the second decade 2007-2017.
```{r, echo=FALSE}
#High Population Deaths 2007-2017
p_high_second <- all_high_second %>%
  group_by(country) %>% 
  ggplot(aes(x= High_Deaths_07, y= High_Deaths_12, color= High_Deaths_17)) +
  geom_point() +
  facet_wrap(~country)

@@

Indoor Deaths
```{r, echo=FALSE}
#Indoor 1997-2017
indoor_high<- deaths_df %>% 
    filter(year> 1995 & country %in% c('United States', 'Brazil','Nigeria', 
                                       'Germany', 'Pakistan', 'Australia')) %>% 
  group_by(country, year, indoor_deaths) %>% 
  ggplot()+
  geom_point(aes(year, indoor_deaths, color = country))+
  ggtitle("High Populated Indoor Death 1997-2017")+
  xlab("Year")+
  ylab("Deaths per 100,000")
ggplotly(indoor_high)

indoor_low<- deaths_df %>% 
    filter(year> 1995 & country %in% c('Canada', 'Chile', 'Malawi', 
                                       'Serbia', 'Sri Lanka', 'Tonga')) %>% 
  group_by(country, year, indoor_deaths) %>% 
  ggplot()+
  geom_point(aes(year, indoor_deaths, color = country))+
  ggtitle("Low Populated Indoor Death 1997-2017")+
  xlab("Year")+
  ylab("Deaths per 100,000")
ggplotly(indoor_low)

#Pakistan and Malawi were both mainly affected in 1996
indoor_top<- deaths_df %>% 
    filter(year> 1995 & country %in% c('Pakistan', 'Malawi')) %>% 
  group_by(country, year, indoor_deaths) %>% 
  ggplot()+
  geom_point(aes(year, indoor_deaths, color = country))+
  ggtitle("Greatest Indoor Death 1997-2017")+
  xlab("Year")+
  ylab("Deaths per 100,000")
ggplotly(indoor_top)
```

@@

Outdoor Deaths
```{r, echo= FALSE}
#Outdoor Particulate 1997-2017
outdoor_high<- deaths_df %>% 
    filter(year> 1995 & country %in% c('United States', 'Brazil','Nigeria', 
                                       'Germany', 'Pakistan', 'Australia')) %>% 
  group_by(country, year, outdoor_deaths) %>% 
  ggplot()+
  geom_point(aes(year, outdoor_deaths, color = country))+
  ggtitle("High Populated Outdoor Death 1997-2017")+
  xlab("Year")+
  ylab("Deaths per 100,000")
ggplotly(outdoor_high)

outdoor_low<- deaths_df %>% 
    filter(year> 1995 & country %in% c('Canada', 'Chile', 'Malawi', 
                                       'Serbia', 'Sri Lanka', 'Tonga')) %>% 
  group_by(country, year, outdoor_deaths) %>% 
  ggplot()+
  geom_point(aes(year, outdoor_deaths, color = country))+
  ggtitle("Low Populated Outdoor Death 1997-2017")+
  xlab("Year")+
  ylab("Deaths per 100,000")
ggplotly(outdoor_low)

#Serbia and Pakistan were the top countries affected. 
#2011 Pakistan had the greatest amount of Outdoor Deaths
outdoor_top<- deaths_df %>% 
    filter(year> 1995 & country %in% c('Pakistan', 'Serbia')) %>% 
  group_by(country, year, outdoor_deaths) %>% 
  ggplot()+
  geom_point(aes(year, outdoor_deaths, color = country))+
  ggtitle("Greatest Outdoor Death 1997-2017")+
  xlab("Year")+
  ylab("Deaths per 100,000")
ggplotly(outdoor_top)
```

***
Ozone Deaths 
```{r, echo=FALSE}
#Outdoor ozone 1997-2017
ozone_high<- deaths_df %>% 
    filter(year> 1995 & country %in% c('United States', 'Brazil','Nigeria', 
                                       'Germany', 'Pakistan', 'Australia')) %>% 
  group_by(country, year, ozone_deaths) %>% 
  ggplot()+
  geom_point(aes(year, ozone_deaths, color = country))+
  ggtitle("High Populated Ozone Death 1997-2017")+
  xlab("Year")+
  ylab("Deaths per 100,000")
ggplotly(ozone_high)

ozone_low<- deaths_df %>% 
    filter(year> 1995 & country %in% c('Canada', 'Chile', 'Malawi', 
                                       'Serbia', 'Sri Lanka', 'Tonga')) %>% 
  group_by(country, year, ozone_deaths) %>% 
  ggplot()+
  geom_point(aes(year, ozone_deaths, color = country))+
  ggtitle("Low Populated Ozone Death 1997-2017")+
  xlab("Year")+
  ylab("Deaths per 100,000")
ggplotly(ozone_low)

#Malawi and Pakistan were the top countries affected 
#1997 had the greatest amount of ozone death in Pakistan
ozone_top<- deaths_df %>% 
    filter(year> 1995 & country %in% c('Pakistan', 'Malawi')) %>% 
  group_by(country, year, ozone_deaths) %>% 
  ggplot()+
  geom_point(aes(year, ozone_deaths, color = country))+
  ggtitle("Greatest Ozone Death 1997-2017")+
  xlab("Year")+
  ylab("Deaths per 100,000")
ggplotly(ozone_top)
```

@@

* Which country has the highest average death count?
  + High Population: Pakistan
  + Low Population: Malawi
* Has the percentage of the affected population decreased or increased over time?
  + Generally it is decreasing for both High and Low populated countries 
* Which pollutant type has the greatest number of deaths?
  + High Population: Indoor Pollution
  + Low Population: Indoor Pollution
* How has the death count changed over the past two decades?
  + 1996-2006: 
    - High Population: Decreases
    - Low Population: Decreases
  + 2007-2017: 
    - High Population: Decreases 
    - Low Population: Decreases
* Which year and country had the highest number of deaths per pollutant type?
  + We looked at years 1996-2017 
  + Indoor: Pakistan and Malawi were mainly affected in 1996
  + Outdoor: Serbia and Pakistan were the top countries. 
    - 2011 was the worst for Pakistan. 
    - 1997 was the worst for Serbia 
    - Sri Lanka and Tonga increased, but Sri Lanka had a steep decrease after 2015 
  + Outdoor Ozone: Pakistan and Malawi were the top countries. 
    - 1997 was the worst for Pakistan. 
    - 1998 was the worst for Malawi
    - United States was the second highest amount of deaths among the higher populated countries
    - Pakistan decreased and then slightly increased
* Which is worse - outdoor or indoor pollution?
  + Inconclusive

# Sources
* [https://www.kaggle.com/datasets/akshat0giri/death-due-to-air-pollution-19902017 ]

* [https://www.worldometers.info/population/world/]

* [https://www.who.int/health-topics/air-pollution#tab=tab_1]

* [https://www.epa.gov/ground-level-ozone-pollution/ground-level-ozone-basics]

* [https://www.health.nsw.gov.au/environment/air/Pages/outdoor-air-pollution.aspx]

* [https://www.mana.md/indoor-air-vs-outdoor-air/]

* [https://www.kaggle.com/datasets/imdevskp/world-population-19602018]

>>>>

Import the deaths-due-to-air-pollution data
```{r}
deaths_df <- data.frame(read.csv("death-rates-from-air-pollution.csv"))
```

@@

```{r, echo=FALSE}
#selecting high-population countries from the world population data frame 
high_pop_countries <- world_pop %>% 
  filter(Year > 1996 & Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
  group_by(Year)

ggplot(high_pop_countries, aes(x=Country.Name, y=Count, color=Country.Name))+
  geom_boxplot()+
  labs(title="Selection of high-population countries", x="Country", y="Population (in millions)")

#selecting low-population countries from the world population data frame 
low_pop_countries <- world_pop %>% 
  filter(Year > 1996 & Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand'))%>% 
  group_by(Year)

ggplot(low_pop_countries, aes(x=Country.Name, y=Count, color=Country.Name))+
  geom_boxplot()+
  labs(title="Selection of low-population countries", x="Country", y="Population (in millions)")

@@

Next, we are going to see the death count for high and low populated countries using the deaths dataframe.
```{r, echo=FALSE}
#selecting high-population deaths from death dataframe 
high_pop_death <- deaths_df %>% 
  filter(year > 1996 & country %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
  group_by(year)

#selecting low-population deaths from death dataframe 
low_pop_death <- deaths_df %>% 
  filter(year > 1996 & country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand')) %>% 
  group_by(year)

@@

# Death Count
> Which country has the highest death count?

@@

```{r, echo=FALSE, message=FALSE}
hp_countries_population <- world_pop %>% 
  filter(Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia'), Year > 1996) %>% 
  group_by(Country.Name) %>% 
  select(Count) %>% 
  summarize(average_population = mean(Count))

lp_countries_population <- world_pop %>% 
  filter(Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand'), Year > 1996) %>% 
  group_by(Country.Name) %>% 
  select(Count) %>% 
  summarize(average_population = mean(Count))

@@

#Low Population Pollutant Averages 
low_poll <- deaths_df %>% 
  group_by(country) %>% 
  filter(country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand')) %>% 
  select(country, indoor_deaths, outdoor_deaths, ozone_deaths) %>% 
  summarize(avg_indoor = mean(indoor_deaths), avg_outdoor = mean(outdoor_deaths), avg_ozone = mean(ozone_deaths))

kable(low_poll, ascending=TRUE)

kable(low_poll)
```

@@

# Pollution Over Time 
> Let's look at the previous two decades and compare the death count
> Has there been a change?

@@

death_06_low<- deaths_df %>% 
  group_by(country) %>% 
  filter(year == 2006, country %in% c('United States', 'Brazil','Nigeria', 'Germany', 'Pakistan', 'Australia')) %>%
  select(total_deaths) %>% 
  summarize(Low_Deaths_06 = mean(total_deaths))

@@

The first decade.
```{r, echo=FALSE}
#High Population Deaths 1996-2006 
p_high_first <- all_high_first %>%
  group_by(country) %>% 
  ggplot(aes(x= High_Deaths_96, y= High_Deaths_01, color= High_Deaths_06)) +
  geom_point() +
  facet_wrap(~country)

@@

This shows the second decade.
```{r, echo=FALSE}
#High Population Deaths 2007-2017
p_high_second <- all_high_second %>%
  group_by(country) %>% 
  ggplot(aes(x= High_Deaths_07, y= High_Deaths_12, color= High_Deaths_17)) +
  geom_point() +
  facet_wrap(~country)

@@

Indoor Deaths
```{r, echo=FALSE}
#Indoor 1996-2017
indoor_high<- deaths_df %>% 
    filter(year> 1995 & country %in% c('United States', 'Brazil','Nigeria', 
                                       'Germany', 'Pakistan', 'Australia')) %>% 
  group_by(country, year, indoor_deaths) %>% 
  ggplot()+
  geom_point(aes(year, indoor_deaths, color = country))+
  ggtitle("High Populated Indoor Death 1996-2017")+
  xlab("Year")+
  ylab("Deaths per 100,000")
ggplotly(indoor_high)

indoor_low<- deaths_df %>% 
    filter(year> 1995 & country %in% c('Canada', 'Chile', 'Malawi', 
                                       'Serbia', 'Sri Lanka', 'Tonga')) %>% 
  group_by(country, year, indoor_deaths) %>% 
  ggplot()+
  geom_point(aes(year, indoor_deaths, color = country))+
  ggtitle("Low Populated Indoor Death 1996-2017")+
  xlab("Year")+
  ylab("Deaths per 100,000")
ggplotly(indoor_low)

#Pakistan and Malawi were both mainly affected in 1996
indoor_top<- deaths_df %>% 
    filter(year> 1995 & country %in% c('Pakistan', 'Malawi')) %>% 
  group_by(country, year, indoor_deaths) %>% 
  ggplot()+
  geom_point(aes(year, indoor_deaths, color = country))+
  ggtitle("Greatest Indoor Death 1996-2017")+
  xlab("Year")+
  ylab("Deaths per 100,000")
ggplotly(indoor_top)
```

@@

Outdoor Deaths
```{r, echo= FALSE}
#Outdoor Particulate 1996-2017
outdoor_high<- deaths_df %>% 
    filter(year> 1995 & country %in% c('United States', 'Brazil','Nigeria', 
                                       'Germany', 'Pakistan', 'Australia')) %>% 
  group_by(country, year, outdoor_deaths) %>% 
  ggplot()+
  geom_point(aes(year, outdoor_deaths, color = country))+
  ggtitle("High Populated Outdoor Death 1996-2017")+
  xlab("Year")+
  ylab("Deaths per 100,000")
ggplotly(outdoor_high)

outdoor_low<- deaths_df %>% 
    filter(year> 1995 & country %in% c('Canada', 'Chile', 'Malawi', 
                                       'Serbia', 'Sri Lanka', 'Tonga')) %>% 
  group_by(country, year, outdoor_deaths) %>% 
  ggplot()+
  geom_point(aes(year, outdoor_deaths, color = country))+
  ggtitle("Low Populated Outdoor Death 1996-2017")+
  xlab("Year")+
  ylab("Deaths per 100,000")
ggplotly(outdoor_low)

#Serbia and Pakistan were the top countries affected. 
#2011 Pakistan had the greatest amount of Outdoor Deaths
outdoor_top<- deaths_df %>% 
    filter(year> 1995 & country %in% c('Pakistan', 'Serbia')) %>% 
  group_by(country, year, outdoor_deaths) %>% 
  ggplot()+
  geom_point(aes(year, outdoor_deaths, color = country))+
  ggtitle("Greatest Outdoor Death 1996-2017")+
  xlab("Year")+
  ylab("Deaths per 100,000")
ggplotly(outdoor_top)
```

***
Ozone Deaths 
```{r, echo=FALSE}
#Outdoor ozone 1996-2017
ozone_high<- deaths_df %>% 
    filter(year> 1995 & country %in% c('United States', 'Brazil','Nigeria', 
                                       'Germany', 'Pakistan', 'Australia')) %>% 
  group_by(country, year, ozone_deaths) %>% 
  ggplot()+
  geom_point(aes(year, ozone_deaths, color = country))+
  ggtitle("High Populated Ozone Death 1996-2017")+
  xlab("Year")+
  ylab("Deaths per 100,000")
ggplotly(ozone_high)

ozone_low<- deaths_df %>% 
    filter(year> 1995 & country %in% c('Canada', 'Chile', 'Malawi', 
                                       'Serbia', 'Sri Lanka', 'Tonga')) %>% 
  group_by(country, year, ozone_deaths) %>% 
  ggplot()+
  geom_point(aes(year, ozone_deaths, color = country))+
  ggtitle("Low Populated Ozone Death 1996-2017")+
  xlab("Year")+
  ylab("Deaths per 100,000")
ggplotly(ozone_low)

#Malawi and Pakistan were the top countries affected 
#1997 had the greatest amount of ozone death in Pakistan
ozone_top<- deaths_df %>% 
    filter(year> 1995 & country %in% c('Pakistan', 'Malawi')) %>% 
  group_by(country, year, ozone_deaths) %>% 
  ggplot()+
  geom_point(aes(year, ozone_deaths, color = country))+
  ggtitle("Greatest Ozone Death 1996-2017")+
  xlab("Year")+
  ylab("Deaths per 100,000")
ggplotly(ozone_top)
```

@@

# Sources
